\documentclass[../main/main.tex]{subfiles}

\newdate{date}{13}{12}{2019}


\begin{document}

\marginpar{ \textbf{Lecture 18.} \\  \displaydate{date}. \\ Compiled:  \today.}

\begin{equation}
  \qty(- \grad ^2 + \xi ^{-2} (t)) G_c (\va{r}-\va{r}') = \frac{k_B T}{k} \delta (\va{r}-\va{r}')
\end{equation}
Let us try to do the Fourier transform.
Let us define
\begin{equation}
  \va{x} \equiv \va{r} - \va{r}'
\end{equation}
e let us call \( \widetilde{G} (q)  \) the Fourier transform of the function \( G \)
\begin{equation}
  \widetilde{G} (q) = \int_{- \infty }^{+ \infty } \dd[]{\abs{\va{x}} } G_c ( \abs{\va{x}} ) e^{- i q \abs{\va{x}} }
\end{equation}
we get
\begin{equation}
  \widetilde{G} ( q ) = \frac{k_B T}{k} \frac{1}{q^2 + \xi ^{-2}}
\end{equation}
At \( T = T_c \), we have  \( \xi \rightarrow \infty  \)  and \( \widetilde{G} (q) \simeq \frac{1}{q^2}  \).
We have
\begin{equation}
  G_c (\abs{\va{x}} ) = \abs{\va{x}}^{2-D}
\end{equation}
In this case we see immediately that \( \eta = 0 \). Go back and find why we have this.
\begin{equation}
  G (\va{x}) = \int_{}^{} \dd[D]{\va{q}}  \frac{1}{(2 \pi )^D} \frac{1}{q^2 + \xi ^{-2}} e^{i \va{q} \vdot \va{x}}
\end{equation}
Let us do it for \( D=3 \):
\begin{equation}
  \Rightarrow G(\abs{x} ) = \frac{4 \pi }{ (2 \pi )^3} \int_{0}^{\infty } \dd[]{q} \frac{q^2}{q^2+ \xi ^{-2}} \int_{-1}^{+1} \dd[]{(\cos \theta  )}    e^{i q \abs{\va{x}} \cos \theta  }
\end{equation}
we get
\begin{equation}
  = \frac{4 \pi }{ (2 \pi )^3} \abs{\va{x}} \int_{0}^{\infty } \dd[]{q} \frac{q \sin(q \abs{\va{x}} ) }{q^2 + \xi ^{-2}}
\end{equation}
At the end
\begin{equation}
  \Rightarrow G ( \abs{\va{x}} ) = \frac{1}{2 \pi } \frac{e^{- \frac{\abs{\va{x}} }{\xi }} }{\abs{\va{x}} }
\end{equation}
Can we reach the simple level of fluctuation? The simple level is the one that follow gaussian distribution.
Let us introduce fluctuations at the Gaussian level.

Idea: consider a function (an usual function)...

Consider \( h=0 \): the saddle point solution is \( m_0 (r) = m_0 \)
\begin{equation}
  \beta \mathcal{H}_{eff} = \int_{}^{} \dd[D]{\va{r}} \qty( a t m^2 + \frac{b}{2} m^4 \frac{k}{2} \qty(\grad m)^2 )
\end{equation}
\begin{equation}
  m(\va{r}) = m_0 + \delta m(\va{r})
\end{equation}
we are assuming that the fluctuations \(  \delta m(\va{r})\) are small.
\begin{equation}
  \qty(\grad m)^2 = \qty(\grad \qty(m_0 + \delta m) )^2 = \qty(\grad \qty(\delta m) )^2
\end{equation}
\begin{equation}
  m^2 = m_0^2 + 2 m_0 \delta m + \qty(\delta m)^2
\end{equation}
\begin{equation}
  \beta \mathcal{H}_{eff} = V \underbrace{ \qty(a t m_0^2 + \frac{b}{2} m_0^4) }_{A_0}
  + \int_{}^{} \dd[D]{\va{r}} \qty(\frac{k}{2}\qty(\grad m)^2 + \qty(a t + 2 b m_0^2)\delta  m^2 + 2 b m_0 \delta m^3 + \frac{b}{2} \delta m^4 )
\end{equation}
\begin{equation}
  \qty( \underbrace{2 a t m_0}_{=0}  + \frac{b}{2} 4 m_0^3 )  \delta m
\end{equation}
the linear term  in \( m_0 \) is equal to zero by definition.

At \( T > T_c \) we know that \( m_0 = 0 \).
In this cases
\begin{equation}
  \beta \mathcal{H}_{eff}^> = \int_{}^{} \dd[D]{\va{r}} \qty(\frac{k}{2} \qty(\grad m)^2 + a t (\delta m)^2 + \cancel{\frac{b}{2} (\delta m)^4} )
\end{equation}
\begin{remark}
It is important to understand that these are fluctuations with respect to the solution.
\end{remark}
We cannot do again the saddle point, otherwise we do not get too much information.
We consider gaussian fluctuations: fluctuations that follow gaussian distribution.
Therefore, the term in \( (\delta m)^4 \)  is cancelled.
What it is the difference in the exponent respect the mean field?
Then we will to the same for \( T < T_c \), the story is the same. Now, we are just taking the gaussian term
\begin{equation}
  Z_{GL}^G = \int_{}^{} \text{D} \qty[\delta m] e^{- \int_{}^{} \dd[D]{r} \qty(\frac{k}{2} \qty(\grad \delta m)^2 + a t (\delta m)^2  )  }
 \end{equation}
Consider a system in a box of volume \( V = L^D \):
\begin{equation}
  \delta m (\va{r}) = \frac{1}{V} \sum_{}^{\va{k}}  e^{i \va{k}\vdot \va{r}} \delta m_{\va{k}}
\end{equation}
We have the integral
\begin{equation}
  \delta m_{\va{k}} = \int_{V}^{} \dd[D]{\va{r}} \delta m (\va{r}) e^{-i \va{k} \vdot \va{r} }
\end{equation}
with \( \va{k} = k_1, \dots, k_D = \frac{2 \pi  \bar{n} }{L}\)
We have \( \delta m_{\va{k}} \in \mathbb{C} \) but \( \delta m (\va{r}) \in \R \), hence
\begin{equation}
  \delta m_{\va{k}} = - \delta m_{-\va{k}}
\end{equation}
\begin{equation}
  \sum_{\va{k}}^{} \rightarrow \frac{V}{( 2 \pi )^D} \int_{\R}^{} \dd[]{\va{k}}
\end{equation}
\begin{equation}
  \frac{1}{V} \sum_{\va{k}}^{} e^{i \va{k} (\va{r}-\va{r}')} \rightarrow \frac{1}{V} \frac{V}{(2 \pi )^2} \int_{\R}^{} \dd[D]{\va{k}} e^{i \va{k} (\va{r}-\va{r}')}
  = \delta (\va{r}- \va{r}')
\end{equation}
\begin{equation}
  \frac{1}{V} \int_{}^{} \dd[D]{\va{r}} e^{i (\va{k} - \va{k}') \vdot  \va{r}} = \delta _{\va{k}\va{k}'}
\end{equation}
write immediately
\begin{equation}
  V \delta _{\va{k} \va{k}'} \overset{V \rightarrow \infty }{\longrightarrow  } (2 \pi )^D \delta (\va{k} - \va{k}')
\end{equation}
\begin{equation}
  a \Rightarrow \abs{\va{k}} \le \frac{\pi }{a} = \Lambda
\end{equation}
that is the ultraviolet cut-off.








\end{document}
